---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we will extract the files from the zip archive, so we can read them in.
The data is a simple comma separated values (csv) formatted table with 3 columns
of information.  We can load all of the data initially into the table using
the read.csv() function.

```{r}
unzip('activity.zip')
activity <- read.csv('activity.csv')
summary(activity)
```

The steps column holds the number of steps that occur in 5 minute intervals,
and we note that there are 2304 missing values for our steps attribute out
of 17568 observations.  Some other things to note, there are 288 observations
each day, because taking readings every 5 minutes over a 24hr * 60min period
results in 288 observations per day.  Knowing there are 288 observations
per day, we can infer recordings were made over a 61 day period, from 
2012-10-01 to 2012-11-31 (because 17568 observations / 288 per day = 61 days)

```{r}
(24 * 60) / 5
17568 / 288
```

The date column is read in as a string, but obviously holds a well formatted
date value.  Likewise, the interval column is read in as an integer, but this
is really representing a time value using a 24 hour clock format.  We will 
probably want to do a lot of processing using the actual date/times, so lets
do a bit of preprocessing.  First of all we will create a datetime attribute 
that uses the date and interval fields to determine a valid datetime object.
And we will also then convert the date column into a true date object.

```{r}
library(lubridate)
activity$datetime <- ymd_hm(paste(activity$date, sprintf("%04d", activity$interval)))
summary(activity$datetime)
activity$date <- ymd(activity$date)
summary(activity$date)
```

## What is mean total number of steps taken per day?

We first want to calculate the total number of steps by day.  There are many
ways to accomplish this in R, but one simple way is to treat the date as
a factor and aggregrate the steps by date using the sum of the steps.

```{r}
stepsperday <- aggregate(steps ~ date, activity, sum) 
stepsperday
```

We note there are only 53 days that have a total steps calculated.  This is
because 8 days in the data set do not have any step data for any interval
during the day (all intervals are NA).  We will deal with this missing
data later.

We can create a histogram of the step totals per day using the basic plotting
system hist() function:

```{r}
hist(stepsperday$steps, breaks=10, main = 'Frequency of total step count per day', xlab = 'Steps')
```

The quartials, as well as the Mean and Median of the total step count per day can
be computed using the summary command.

```{r}
summary(stepsperday$steps)
```

## What is the average daily activity pattern?

Here we will do basically the opposite of our previous summary and aggregrate
the steps activity using the interval as our factor.  Here we want to aggregrate
the average number of steps in that interval.  

```{r}
stepsperinterval <- aggregate(steps ~ interval, activity, mean) 
```

Now plot this data as a time series, using the interval on the x axis, and
our mean activity data for the y axis.

```{r}
with(stepsperinterval, plot(interval, steps, type='l', main = 'Average steps per interval', xlab='time of day'))
```

From the figure we can see that for this data sometime in the early morning,
before 10:00, is when the maximum average number of steps occur.  This
maximum occurs at:

```{r}
stepsperinterval[stepsperinterval$steps == max(stepsperinterval$steps),]
```

So we see for this data the maximum is occuring for the 8:35 interval, and
averages just over 206 steps per day during this period.

## Imputing missing values

The total number of missing values for the steps variable is 2304 missing
values.

```{r}
summary(activity$steps)
sum(is.na(activity$steps))
```

One question not asked in this assignment is, do the NA values represent
days that are completely missing, or are there some days that partially
have step values, and partially have some NA values.  I believe it is
probably the former.  As we have noted from previous work, there are
a total of 8 days that completely have NA missing values.  8 days
times 288 missing values

```{r}
8 * 288
```

is exactly 2304, which seems likely not to be a coincedence.  One way to check
is to extract only those observations with NA values for the steps, then
see which set of dates NA values occur on:

```{r}
missing <- activity[is.na(activity$steps),]
levels(as.factor(missing$date))
```

As can be seen there are only 8 days that contain missing data.  Also these
correspond to those 8 days that were missing when we summed the steps by day.

Knowing that it is basically 8 whole days that are missing activity data, it 
makes the most sense that we represent these missing days by using
the average steps by interval to fill them in.  We already have an example
of calculating this aggregration when we created our time series above.
One further question, since these are averages they will no longer be
whole integer units.  I suppose there might be some analysis where these
not being whole numbers as we would expect might cause a problem in which
case we could simple round these averages to their nearest values to use
as our approximation.  I have not done this, but the round() function could
easily be added to the next bit of code where we imute our missing values.

So the idea is that for each day we want to impute the mean value for
the given interval to replace each NA value.  If we are completely sure that
whole days are missing, the simplest method is to use R's ability to
repeat values when assigning one list/arrays values to a larger set.  We
can impute our 8 missing days, then, in the following manner:

```{r}
missing.index <- is.na(activity$steps)
stepsperinterval <- aggregate(steps ~ interval, activity, mean) 
activity[missing.index, 'steps'] <- stepsperinterval$steps
sum(is.na(activity$steps))
```

This successfully replaces all of the NA missing values.  Because there were
288 mean values in the stepsperinterval$steps data, this set of values
was repeated 8 times to fill in the 2304 missing NA steps values.

So now we redo our calculation of the total number of steps per day
calculation, and display a histogram of the data.

```{r}
stepsperday <- aggregate(steps ~ date, activity, sum) 
hist(stepsperday$steps, breaks=10, main = 'Frequency of total step count per day', xlab = 'Steps')
```

Basicaly as you would expect, we ended up adding 8 days that now have exactly the
average total number of steps, so the largest bin has gone from a frequency of
16 to 24 now for this histogram.  Looking at the mean and median:

```{r}
summary(stepsperday$steps)
```

The mean value has not changed, as we would expect since we added 8 days with
the same mean as the original 53 nonmissing days.  However, the median value
does change slightly, so that mean and median are basically about the same now.


## Are there differences in activity patterns between weekdays and weekends?

Lets first create a factor variable to separate the days into those on the
weekend (Saturday or Sunday) or those during the weekday.  

```{r}
activity$daytype <- 'weekday'
weekend.index <- weekdays(activity$date) == "Saturday" | weekdays(activity$date) == "Sunday"
activity[weekend.index, 'daytype'] = 'weekend'
activity$daytype <- as.factor(activity$daytype)
summary(activity$daytype)
```

Now we will create a panel plot of time series data comparing the average 
number of steps by interval, breaking out on the weekday/weekend factor.

```{r}
s <- aggregate(steps ~ interval + daytype, activity, mean)
weekday = s[s$daytype == 'weekday',]
weekend = s[s$daytype == 'weekend',]
```

```{r}
par(mfrow = c(2,1))

with(weekday, plot(interval, steps, type='l', main = 'Weekday activity'))
with(weekend, plot(interval, steps, type='l', main = 'Weekend activity'))
```

```{r}
library(lattice)
xyplot(steps ~ interval | daytype, data = s, type = 'l', layout=c(1,2))
```
